Parallel Adaptive Importance Sampling

### 2016-06-02 PAIS.tex

Instead of sampling $M$ times from an $M$-dimensional multinomial
distribution as is the case with the bootstrap algorithm, we sample
once each from $M$ different multinomials. Suppose that we have $M$
samples $y_n$ with weights $w_n$. The multinomial sampled from in the
bootstrap filter has a vector of probabilities given by:
\begin{equation*}
\frac{1}{\sum w_n} [w_1,w_2,\ldots,w_M] = \bar{\bf w},
\end{equation*}
with associated states $y_n$.
We wish to find $M$ vectors $\{{\bf p}_1,{\bf p}_2,\ldots,{\bf p}_M\}
\subset \mathbb{R}^M_{\ge 0}$
such that  $\frac{1}{M} \sum {\bf p}_i = \bar{\bf w}$. The AMR is then
given by a sample from each of the multinomials defined by the vectors
${\bf p}_i = [p_{i,1},p_{i,2},\ldots,p_{i,M}]$ with associated states ${\bf y}_i$. Alternatively, as with the ETPF, a deterministic sample
can be chosen by picking each sample to be equal to the mean value of
each of these multinomial distributions, i.e. each new sample
$\hat{x}_i$ is given by:
\begin{equation}
\hat{x}_i = \sum p_{i,j} x_j, \qquad i \in \{1,2,\ldots,M\}.
\end{equation}

The resulting sample has several properties which are advantageous in
the context of being used with the PAIS algorithm. Firstly, we have
effectively chopped up the multinomial distribution used in the
bootstrap filter into $M$ pieces, and we can guarantee that exactly
one sample will be taken from each section. This leads to a much
smaller chance of losing entire modes in the density, if each of the
sub-multinomials is picked in an appropriate fashion. Secondly, if we do not make a random sample for
each multinomial with probability vector $\bf p_i$ but instead take
the mean of the multinomial to be the sample, this algorithm preserves
the mean of the sample exactly. Lastly, as we will see shortly, this
algorithm is significantly less computationally intensive than the
ETPF.

@@

Table \ref{tab:AMR} describes the basis of the algorithm with
deterministic resampling, using the means of each of the
sub-multinomials as the new samples. This resampler was designed with the aims of being
more efficient than the ETPF, and more accurate than straight multinomial
resampling. Therefore we now present numerical examples which
demonstrate this.

@@

To test the accuracy and speed of the three resamplers (ETPF,
bootstrap and AMR), we drew a sample of size $M$ from the proposal distribution
$\mathcal{N}(1,2)$. Importance weights were assigned, based on a
target distribution of $\mathcal{N}(2,3)$. The statistics of the
resampled outputs were compared with the original weighted samples. Figure \ref{fig:AMR} (a)-(c) show how the relative errors in the first
three moments of the samples changes with $M$ for the three different
samplers. As expected, the AMR lies somewhere between the high
accuracy of the ETPF and the less accurate bootstrap
resampling. Note that only the error for the bootstrap multinomial
sampler is presented for the first moment since both the ETPF and the
AMR preserve the mean of the original weighted samples up to machine precision. Figure \ref{fig:AMR} (d) shows how the computational cost,
measured in seconds, scales with the ensemble size for the three
different methods. These results demonstrate that the AMR behaves how we wish, and
importantly ensures that exactly one sample of the output will lie in
each region with weights up to $\frac{1}{M}$ of the total.

>>>>

Instead of sampling $M$ times from an $M$-dimensional multinomial
distribution as is the case with the bootstrap algorithm, we sample
once each from $M$ different multinomials. Suppose that we have $M$
samples $y_n$ with weights $w_n$. The multinomial sampled from in the
bootstrap filter has a vector of probabilities given by:
\begin{equation*}
\frac{1}{\sum w_n} [w_1,w_2,\ldots,w_M] = \bar{\bf w},
\end{equation*}
with associated states $y_n$.
We wish to find $M$ vectors $\{{\bf p}_1,{\bf p}_2,\ldots,{\bf p}_M\}
\subset \mathbb{R}^M_{\ge 0}$
such that  $\frac{1}{M} \sum {\bf p}_i = \bar{\bf w}$. The AMR is then
given by a sample from each of the multinomials defined by the vectors
${\bf p}_i = [p_{i,1},p_{i,2},\ldots,p_{i,M}]$ with associated states ${\bf x}_i$. Alternatively, as with the ETPF, a deterministic sample
can be chosen by picking each sample to be equal to the mean value of
each of these multinomial distributions, i.e. each new sample
$\hat{x}_i$ is given by:
\begin{equation}
\hat{x}_i = \sum p_{i,j} x_j, \qquad i \in \{1,2,\ldots,M\}.
\end{equation}

The resulting sample has several properties which are advantageous in
the context of being used with the PAIS algorithm. Firstly, we have
effectively chopped up the multinomial distribution into used in the
bootstrap filter into $M$ pieces, and we can guarantee that exactly
one sample will be taken from each section. This leads to a much
smaller chance of losing entire modes in the density. Secondly, if we do not make a random sample for
each multinomial with probability vector $\bf p_i$ but instead take
the mean of the multinomial to be the sample, this algorithm preserves
the mean of the sample exactly. Lastly, as we will see shortly, this
algorithm is significantly less computationally intensive than the
ETPF.

@@

Table \ref{tab:AMR} describes the basis of the algorithm with
deterministic resampling, using the means of each of the
sub-multinomials. This resampler was designed with the aims of being
more efficient than the ETPF, and more accurate than straight multinomial
resampling. Therefore we now present numerical examples which
demonstrate this.

@@

To test the accuracy and speed of the three resamplers (ETPF,
bootstrap and AMR), we drew a sample of size $M$ from the proposal distribution
$\mathcal{N}(1,2)$. Importance weights were assigned, based on a
target distribution of $\mathcal{N}(2,3)$. The statistics of the
resampled outputes were compared with the original weighted samples. Figure \ref{fig:AMR} (a)-(c) show how the relative errors in the first
three moments of the samples changes with $M$ for the three different
samplers. As expected, the AMR lies somewhere between the high
accuracy of the ETPF and the less accurate bootstrap
resampling. Note that only the error for the bootstrap multinomial
sampler is presented for the first moment since both the ETPF and the
AMR preserve the mean up to machine precision. Figure \ref{fig:AMR} (d) shows how the computational cost,
measured in seconds, scales with the ensemble size for the three
different methods. These results demonstrate that the AMR behaves how we wish, and
importantly ensures that exactly one sample of the output will lie in
each region with weights up to $\frac{1}{M}$ of the total.
